Building the Million Transcriptome Ocular Meta-Atlas

### 2021-01-16 manuscript_00.Rmd
conclusion first draft

The four genes with zero hits were SCRG1 (Astrocytes), BTBD8 (Bipolar Cells), GATM (Microglia), and FRZB (Muller Glia). In a more exhaustive literature search we find that SCRG1 is a secreted protein which marks brain neurons, regulates human mesenchymal stem cells differentiation, and is overexpressed in brain astrocyte precursors [@aomatsuNovelSCRG1BST12014; @chaboubTemporalProfilingAstrocyte2016]. BTBD8 contains BTB/POZ domains known to generally be associated with with transcription factor mediated protein-protein interactions [@xuMolecularCloningCharacterization2004; @zollmanBTBDomainFound1994]. GATM is mitochondriale nzyme involved in creatine biosynthesis and was previously shown to upregulated in a brain microglia population [@hammondSingleCellRNASequencing2019]. FRZB was previously referred to as SFRP3 and this Wnt pathway member has been previously shown to be active in Muller Glia [@zhuCharacterizationCanonicalWnt2018]. Overall, we find very high concordance with known retina cell type markers.

@@

## Projection of new data onto scEiaD

The hard work of creating this resource can be leveraged and extended by the wider community with a few relatively simple steps and modest compute requirements. Very briefly, if outside groups quantify their mouse or human scRNA with kallisto and the same references (see methods), they can overlay their data on top of scEiaD by 1. installing scVI >=0.8.0, 2. downloading our ~13 megabyte scVI model, and 3. following the Jupyter notebook on Google colab that we provide as supplemental file BLOOP and at http://colab.google. We demonstrate the power of this approach in two ways.

First, we project another retina dataset onto scEiaD from Fadl et al.. As this is a pan-retina tissue sample, we would expect most of the major cell types of the retina to be present. In the Google colab notebook we provide a short set of the reads from SRR12130660 can be processed from the raw reads to a UMAP overlapping scEiaD in under 15 minutes. Indeed, as expected we see many amacrine, rods, muller glia, bipolar, a few cones and retinal ganglion, and a small number of non-retina cells (`r supFig_cap(name = 'projection_fig', display = 'cite')`b).

Second, we demonstrate how the RPE derived from the Bharti iPSC differentiation method closely match tissue RPE from other studies (the largest being the Cowan et al data; `r supFig_cap(name = 'projection_fig', display = 'cite')`a) This provides another layer of evidence that the Bharti iPSC-RPE differentation protocol generates RPE cells that are functionally equivalent to human adult RPE.

@@

scran findmarkers (wilcox test) along with singleCellHaystack. Filtering to keep genes that mean AUC > 0.2 or with a D KL > 0.2.

@@

## Projection of new data onto scEiaD

@@

## Self built meta-atlases are easy as 1,2, benchmark a bunch, 3

@@

## The scEiaD dataset provides a highly diverse, large N resource for methods development

## The scEiaD is a unique ocular community resource

@@

>>>>

The four genes with zero hits were SCRG1 (Astrocytes), BTBD8 (Bipolar Cells), GATM (Microglia), and FRZB (Muller Glia). In a more exhaustive literature search we find that SCRG1 is a secreted protein which marks brain neurons, regulates human mesenchymal stem cells differentiation, and is overexpressed in brain astrocyte precursors [@aomatsuNovelSCRG1BST12014; @chaboubTemporalProfilingAstrocyte2016]. BTBD8 contains BTB/POZ domains known to generally be associated with with transcription factor mediated protein-protein interactions [@xuMolecularCloningCharacterization2004; @zollmanBTBDomainFound1994]. GATM is mitochondrial enzyme involved in creatine biosynthesis and was previously shown to up-regulated in a brain microglia population [@hammondSingleCellRNASequencing2019]. FRZB was previously referred to as SFRP3 and this Wnt pathway member has been previously shown to be active in Muller Glia [@zhuCharacterizationCanonicalWnt2018]. Overall, we find very high concordance with known retina cell type markers.

@@

## Projection of new data onto scEiaD demonstrates how iPSC-based cells are similar to tissue-based cells

The hard work of creating this resource can be leveraged and extended by the wider community with a few relatively simple steps and modest compute requirements. Very briefly, if outside groups quantify their mouse or human scRNA with kallisto and the same references (see methods), they can overlay their data on top of scEiaD by 1. installing scVI >=0.8.1, 2. downloading our 13 megabyte scVI model, and 3. following the Jupyter notebook on Google colab that we provide as supplemental file BLOOP and as a live demo at http://colab.google. We demonstrate the power of this approach in two ways.

First, we project another retina dataset onto scEiaD from Fadl et al . As this is a pan-retina tissue sample, we would expect most of the major cell types of the retina to be present. In the Google colab notebook we show how a set of the reads from SRR12130660 can be processed from the raw reads to a UMAP visualization overlapping our scEiaD in under 15 minutes. Indeed, as expected we see many amacrine, rods, muller glia, bipolar, a few cones and retinal ganglion, and a small number of non-retina cells (`r supFig_cap(name = 'projection_fig', display = 'cite')`b).

Second, we demonstrate how the RPE derived from the Bharti iPSC differentiation method closely match tissue RPE from other studies (the largest being the Cowan et al data; `r supFig_cap(name = 'projection_fig', display = 'cite')`a). Cells generated with this process have been extensively validated with functional RPE tests. Our scRNA approach provides another layer of evidence that the Bharti iPSC-RPE differentation protocol generates RPE cells that are functionally equivalent to human adult RPE in an unbiased manner.

@@

To identify marker genes across the CellType (predict) and cluster groups, we used the scran findmarkers (wilcox test) along with the singleCellHaystack algorithm. The scran findmarkers test runs a wilcox test in a pairwise manner (e.g. Rods vs all other cell types). It returns an overall p-value (and FDR) that assesses how well the gene is at separating the group of interest from all other cells. It also returns for each pair-wise comparison an area under the curve (AUC) score, where 1 is a perfect power to distinguish and 0 is no power. The singleCellHaystack uses a Kullback-Leibler divergence measurement of the scVI lower dimensional space to identify genes with non-random distribution. A higher "D KL" score represents a gene with "specific" expression in teh UMAP/lower dimensional space. We filtered to keep genes with FDR < 0.01 and a mean AUC > 0.2 or  a D KL > 0.2.

@@

@@

# Limitations

The scVI model is first built the on human data. The mouse and macaque data are then projected (or queried) onto it with the scVI/scArches approach. While this system works very well to integrate information between these three species, this approach may not scale to more distantly related species. Another limitations is that discrepancies between cell type labels between different labs makes certain transitioning cell type labels a bit imprecise. One example is how the rods and photoreceptor precursor labels overlap somewhat. Though we attempted to ameliorate the issue by removing cell type labels in large disagreement with the consensus our cell type labels are effectively "crowd-sourced" and some disagreements can propagate into our labeled cells.

## The scEiaD is a unique ocular resource that provides a highly diverse, large N dataset with a relatively small amount of compute power

We have assembled the largest ocular single cell transcriptome database to date. The rapid of advancement of algorithms to batch correct and process data continue to reduce the computational requirements to handle huge numbers of cells. The scVI batch correction step on around one million cells runs within a few hours on a GPU and 150GB of memory. This places this crucial step within the capabilities of a moderately large physical server or a cloud compute node. Further downstream proessing can largely be done a computers with 64+ Gb of memory and a few hundred GB of disk space. We believe that our efforts can be replicated in any other tissue / system with a large number of independent studies by an individual lab or small number of scientists. We provide the completed analysis as both Scanpy (h5ad) and Seurat objects at https://github.com/davemcg/scEiaD.

## Benchmarking and quantitation of integration performance is crucial for meta-atlas studies

We originally intended to just use the Seurat CCA method to integrate the datasets. However, the long run-times of CCA and poor integration of our known cell types led us to benchmark more methods and parameters. We also originally attempted to "hand-assess" the integration results by using the UMAP 2D projection view. This proved to scale poorly and this led use to curate some of the most useful benchmarking algorithms (NMI, ARI, silhouette, and LISI) that roughly matched our "hand-assessed" results into the scPOP R package. While we chose the scVI algorithm, we do not believe that it will work best for all datasets. We strongly suggest any other groups attempting a similar meta-atlas construction chose a quantifiable set of criteria so optimal methods and parameters can be picked.

@@

We hand curate over 300,000 cell type labels. With a standard xgboost algorithm using the latent dimensions, cell age classification (developing or matured), and the UMAP coordinates, we can very accurately label the remaining cells. Many other cell type labelling algorithms and systems exist for those groups less willing or able to tune a machine learning algorithm. For example, the developers of scVI also have a cell type label projection algorithm called scANVI. Whichever approach you use, taking a smaller number of high quality labels and projecting them onto the remaining cells is a powerful way to leverage community knowledge across a huge diverse dataset.

@@

Many retina "atlases" have been published to date. We argue that we have created the first atlas that is generally useful because 1. our dataset/atlas is several times larger than any other published set, 2. our data is available via download in several forms at https://github.com/davemcg/scEiaD, and crucially 3. we provide a Google colab/Jupyter notebook which exactly lays out how to use scVI's to project (or query) outside data onto our scEiaD with minimum compute resources. Finally we also provide our xgboost model so outside groups can label their own datasets with our robust retina cell type labelling system.
